Last glacial cycle glacier erosion potential in the Alps

### 2021-05-18 articles/erosion/response-referee2.md

## General Comments

> The authors provide large amounts of very useful results from simulations,
> concerning the effects of varying glacier geometry on erosion potential. It
> was quite an achievement to compute this at fairly high temporal and spatial
> resolution (1 km and 50 a throughout the Alps). The conclusions are
> reasonable, and I value the results. They provide support for the
> traditional view, that high-altitude features such as cirques are eroded by
> steep, local glaciers and not under extensive icefields well above ELA. Most
> erosion is expected along major valleys (glacial troughs): the potential
> erosion is greatest low down the glaciers, with a lesser maximum at high
> altitude and a minimum between. The paper is well written and the Figures
> are all relevant and interesting. The qualitative conclusions are well
> justified.

> An interesting result, not commented on, is that from Figs. 3 and 7 i & j,
> potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm
> sea level equivalent. I would love to see a map of the glaciers at that
> stage\*, which is well below maximum but considerably more than recent ice
> cover (0.3 ± 0.1 mm s.l.e., Farinotti et al. 2019). This peak applies to the
> ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws
> which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).

Thank you for pointing this out. We are not confident that this result is
significant. The situation you describe occurs, for instance, during much of
MIS~5 (Fig.~2) and can be visualized in the animations. During such periods,
glaciers are steep and only a few grid cells in width. Their velocities and
thus computed erosion rates may be overestimated by the limiting horizontal
resolution and the shallow-ice physics (Imhof et al., 2019).

Increasing ice volume yields the build-up of less steep and thus slower-flowing
valley glaciers. Decreasing ice volume yields to (roughly) equally steep but
smaller glaciers. In the case of non-linear erosion laws ('Koppes' and 'Herman'
laws), both yield a decrease of erosion volume, hence the local peak in erosion
volume. We now emphasize this uncertainty in several of our plots by using
hatches over areas of limited glacier cover, documented in the figure captions.

### 2021-05-19 articles/erosion/response-referee1.md

> I think the study could also well contribute to the discussion on the
> elevation distribution of the cumulative glacial erosion over several cycles
> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
> plot showing the integrated glacial erosion potential over (e.g. 100 m)
> elevation bins would easy to do and eventually an enlightening supplement
> summing up Fig. 5a.

TODO: try and compute cumulative erosion by elevation bins.

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

TODO: I think this will best fit in "age of the glacial landscape".

@@

In several instances we replaced "yet" with "but" and "however".

## Figures

> Fig. 2 Please indicate the outline of the connected ice stream network during
> the LGM (like in Fig. 4 of Seguinot et al., 2018). Especially in the eastern
> and SW Alps coverage largely deviates from what has been suggested from field
> data compilation (e.g. Ehlers and Gibbard, 2004). Even though outcomes might
> not be changing much, it might be helpful to know how much the %overlap is -
> erosional potential might change as fluvial topography turns into glacial one
> (e.g. Harbor, 1988) and this is probably not what you want to mix, I guess.

The LGM outline was added on Fig. 2.

TODO: discuss shortly the discrepancy in "climate sensitivity".
TODO: I don't understand the second part of the comment.

@@

> Sanders, J.W., Cuffey, K.M., Moore, J.R., MacGregor, K.R., and Kavanaugh,
> J.L., 2012. Periglacial weathering and headwall erosion in cirque glacier
> bergschrunds: Geology, v. 40, p. 779-782.

> Sternai, P., Herman, F., Valla, P.G., Champagnac, J.-D., 2013. Spatial and
> temporal variations of glacial erosion in the Rhône valley (Swiss Alps):
> Insights from numerical modeling. Earth and Planetary Science Letters, 368,
> p. 119-131.

> Valla, P.G., Shuster, D.L., van der Beek, P.A., 2011. Significant increase in
> relief of the European Alps during mid-Pleistocene glaciations. Nat. Geosci.
> 4, 688–692.

TODO: I am not sure what to do about this long list of papers. Some of them are
already cited, some just added in.

### 2021-05-19 articles/erosion/alps-erosion.tex
Add notes about reviewer #1 generic comments.
Begin work on reviewer #2 generic comments.
Add LGM outline on cumulative erosion figure.

    \begin{figure*}
      \centerline{\includegraphics{alpero_cumulative}}
      \caption{%
        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
          potential over the last glacial cycle.
        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
          equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
          volume (light brown) and 1-ka running mean (dark brown). Shaded gray
          areas indicate the timing for MIS~2 and~4
          \citep{Lisiecki.Raymo.2005}.
        \todo{add LGM outline}}
        \label{fig:cumulative}
    \end{figure*}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Modelled erosion rate ``hypsogram'', showing the geometric
          mean of (non-zero) modelled erosion rates in 10-m elevation bands
          across the entire model domain and its time evolution. The histogram
          to the side shows the distribution of topography in the model domain.
        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.
      \todo{test cutting the hypsogram below a minimum number of grid cells}}
      \label{fig:hypsogram}
    \end{figure*}

>>>>

    \begin{figure*}
      \centerline{\includegraphics{alpero_cumulative}}
      \caption{%
        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
          potential over the last glacial cycle and geomorphological
          reconstruction of Last Glacial Maximum Alpine glacier extent for
          comparison \citep[solid red line,][]{Ehlers.etal.2011}.
          The background maps consists of the initial basal topography from
          SRTM \citep{Jarvis.etal.2008} and Natural Earth Data
          \citep{Patterson.Kelso.2017}.
        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
          equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
          volume (light brown) and 1-ka running mean (dark brown). Shaded gray
          areas indicate the timing for MIS~2 and~4
          \citep{Lisiecki.Raymo.2005}.
        \todo{hatches on all plots.}}
        \label{fig:cumulative}
    \end{figure*}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpero_hypsogram}}
      \caption{%
        \textbf{(a)} Modelled erosion rate ``hypsogram'', showing the geometric
          mean of (non-zero) modelled erosion rates in 10-m elevation bands
          across the entire model domain and its time evolution. The histogram
          to the side shows the distribution of topography in the model domain.
        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.
      \todo{%
        test cutting the hypsogram below a minimum number of grid cells,
        try and compute cumulative erosion by elevation bins.}}
      \label{fig:hypsogram}
    \end{figure*}
